---
title: "CHM2Segment workflow"
author: "Andrii Zaiats"
date: "Last updated: 2022-09-06"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE, echo = TRUE, eval = FALSE)
```

#### Canopy Height Model (CHM) to segmented crowns workflow using _lidR_ package.\

The script goes step by step from the canopy height model to a final product that is a geospatial layer of segmented crowns. The following code is a test version of the workflow and may undergo changes/improvements.\

##### *Steps in the workflow:*

##### *Required software*

-   R packages: *lidR, future, sf, terra, stars, tidyverse, *.

Setting up the environment: first we load the R packages, then provide system paths to the input file(s) and where the output should be stored. The three paths are: (1) path to the location of the dense cloud [*Project folder* above], (2) path to the *tiles* folder, and (3) path to the *test* folder. Lastly, we can specify site name, date of data collection, and CRS for consistent naming convention.

```{r, results='hide', message=FALSE}
# load packages
pkgs <- c("lidR", "raster", "gstat", "sf", "sp", "viridisLite", "stars", "dplyr")
sapply(pkgs, require, character.only = TRUE, quietly = TRUE)


# specify paths
path_prj <- "D:/Andrii/20220524_NorthHam/"
path_tiles <- paste0(path_prj, "tiles/")
path_test <- paste0(path_prj, "test/")


# data attributes
site <- "NorthHam"
date <- "20220524" # in YYYYMMDD format
prj_crs <- "wgs84utm11N"
```
